---
title: Nürnberg - 3. Greenness Accessibility Index (GACI)
author: ''
date: '2025-01-15'
slug: nuernberg-greenness-accessibility-index
categories: ["R"]
tags: ["Nürnberg", "GACI", "R", "Greenness Accessibility", "Greenspace", "CGEI"]
summary: "By integrating realistic walking routes and park size weighting, the Greenspace Accessibility Index (GACI) offers a more practical and community-focused measure of how easily residents can reach quality public parks."
output: blogdown::html_page
image:
  focal_point: ""
  placement: 2
  preview_only: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The **accessibility of greenspace** is a critical factor for urban health and well-being, focusing on how easily people can physically or legally reach parks, community gardens, or forests. Scholars typically assess accessibility by measuring walking or driving distance to the nearest park - a process that can rely on either **Euclidean** or **road network** analyses. However, information on structural or legal access to these areas is not always readily available, particularly at larger geographic scales [@labib2020_review]. Despite these challenges, greenspace accessibility is highly relevant for both **instoration** (building capacities) and **restoration** (restoring capacities) [@markevych2017].

Research indicates that **access to green areas** can foster **social cohesion**, described as a sense of belonging and mutual respect among neighbors [@holtan2014; @weinstein2015]. Greater social cohesion has in turn been linked to **improved mental health outcomes** and enhanced well-being [@fone2014; @williams2020], potentially moderating the impacts of stressful life events [@kingsbury2019]. Alongside these social benefits, accessible greenspaces also promote **physical activity** - an important tool for managing mental health conditions and boosting overall health [@lachowycz2011; @duncan2014; @mcgrath2015]. Studies suggest that “green exercise” yields higher psychological gains than similar activities in less vegetated environments [@pretty2005; @mitchell2013].

Nonetheless, not all greenspaces are equally beneficial for social interaction or physical activity. A **large public park** with amenities to support community gatherings, for instance, may offer different health benefits compared to a **small patch of roadside greenery** [@markevych2017]. Therefore, refined measures of greenspace accessibility - such as location-specific amenities, park size, and the presence of safe walkable routes - are needed to understand how and why some spaces provide stronger instoration and restoration benefits than others [@giles-corti2005].

## Data

The relevant spatial datasets have been uploaded on GitHub: administrative boundaries (AOI), parks and water mask shapefiles.

```{r warning=FALSE, message=FALSE}
library(sf)
library(dplyr)
library(terra)
library(CGEI)

data_repo <- "https://github.com/STBrinkmann/SpatData_Nbg/raw/refs/heads/master/01_analysis/0102_data/02_processed/"

# AOI
aoi <- read_sf(paste0(data_repo, "01_Nbg_Bezirke.gpkg")) %>%
  filter(code_bez != "97") %>% # This is not available in the GAVI
  group_by(code_bez, name_bez) %>%
  summarise() %>%
  ungroup()

# Water mask
water_mask <- read_sf(paste0(data_repo, "02_Nbg_Water.gpkg"))

# Parks (>1ha)
osm_parks <- read_sf(paste0(data_repo, "02_Nbg_Greenspace.gpkg")) %>%
  mutate(name = stringr::str_pad(1:n(), 3, pad = "0")) %>%
  mutate(size_ha = st_area(geom) %>%
           units::set_units(value = ha) %>%
           as.numeric()) %>%
  filter(size_ha > 1) %>%
  mutate(size_log = log(size_ha)) %>%
  relocate(geom, .after = last_col())
```

```{r echo=FALSE, fig.height=8.5, fig.width=10, message=FALSE, warning=FALSE}
library(mapsf)

moegeldorf <- aoi[aoi$code_bez == "92",]

par(bg=NA)
mf_theme(bg = "transparent")

# 1. Main map
mf_map(aoi, expandBB = c(0, 0, 0, 0))
mf_map(st_intersection(osm_parks, aoi), add = TRUE, col = "green4")
mf_map(st_intersection(water_mask, aoi), add = TRUE, col = "darkblue", border = "darkblue")

# 3. Credits and scale bar
mf_credits(txt = paste("Sources", 
                           "OpenStreetMap", 
                           "www.statistik.nuernberg.de", 
                           "Own calculations", 
                           sep = "\n"),
           cex = 0.8)
mf_scale(size = 5, cex = 0.8, pos = c(653880, 5466356))
```

## GACI

The **Greenspace Accessibility Index (GACI)** [@labib2021_greenspaceExposure] quantifies the ease with which the public can access **urban parks**. Rather than relying solely on Euclidean distance, GACI explicitly incorporates:

1.  **Park Access Points**\
    Identifiable entry sites - typically at the intersection of park boundaries with streets or footpaths - where users can legally and physically enter a park.

2.  **Network-Based Distance**\
    Walking distance is determined through road and footpath networks, capturing realistic travel times rather than simple straight-line distances.

3.  **Park Size Weighting**\
    Larger parks potentially attract users from a greater distance. Consequently, GACI considers both distance and the (log-transformed) size of the park.

4.  **Normalization and Classification**\
    Extreme values (e.g., very large distances) are trimmed to mitigate outliers, then final scores are **reclassified** - often via **Jenks natural breaks** - to yield an intuitive map of greenspace accessibility.

Together, these factors yield a spatial indicator that aligns more closely with **real-world access** than conventional 2D buffer analyses.

### 1. Identifying Park Access Points

Public parks are generally entered where roads or paths meet park boundaries. We **intersect** an OpenStreetMap (OSM)-based park polygon layer with an OSM road network. Where no intersection is found, park **centroids** serve as fallback access points. This step ensures each park has at least one feasible entry location.

```{r eval=FALSE}
# --- 1. Generate Road Network & Park Access Points ---

# A. Create buffered AOI boundary to mitigate edge effects
aoi_bbox <- aoi %>%
  st_buffer(5000) %>%
  st_transform(4326) %>%
  st_bbox()

# B. Download and filter OSM road network for walkable paths
osm_roads_raw <- opq(bbox = aoi_bbox) %>%
  add_osm_feature(key = "highway") %>%
  osmdata_sf(quiet = FALSE) %>%
  osm_poly2line()

osm_roads <- osm_roads_raw$osm_lines %>%
  select(highway) %>%
  filter(!highway %in% c("motorway", "motorway_link", "trunk", "trunk_link", "raceway")) %>%
  st_transform(st_crs(aoi)) %>%
  st_intersection(st_buffer(aoi, 5000))

# C. Segment roads for higher intersection accuracy
cores <- 12
if (Sys.info()[["sysname"]] == "Windows") {
  cl <- parallel::makeCluster(cores)
  osm_roads <- suppressWarnings(split(osm_roads, seq(from = 1, to = nrow(osm_roads), by = 200)))
  osm_roads <- parallel::parLapply(cl, osm_roads, function(x){
    nngeo::st_segments(sf::st_cast(x, "LINESTRING"), progress = FALSE)
  })
  parallel::stopCluster(cl)
} else {
  osm_roads <- suppressWarnings(split(osm_roads, seq(from = 1, to = nrow(osm_roads), by = 200))) %>%
    parallel::mclapply(function(x){
      nngeo::st_segments(sf::st_cast(x, "LINESTRING"), progress = FALSE)
    },
    mc.cores = cores, mc.preschedule = TRUE)
}

osm_roads <- st_as_sf(dplyr::as_tibble(data.table::rbindlist(osm_roads)))
osm_roads <- st_set_geometry(osm_roads, "geom")

st_geometry(osm_roads) <- st_geometry(osm_roads) %>%
  lapply(function(x) round(x, 0)) %>%
  st_sfc(crs = st_crs(osm_roads))

# D. Derive park centroids and road intersections
osm_park_centroids <- osm_parks %>%
  st_centroid() %>%
  select(name, size_ha, size_log)

# Helper function
st_multipoint_to_point <- function(x) {
  x.points <- x[which(sf::st_geometry_type(x) == "POINT"), ]
  x.multipoints <- x[which(sf::st_geometry_type(x) == "MULTIPOINT"), ]

  for (i in 1:nrow(x.multipoints)) {
    x.points <- x.points %>%
      dplyr::add_row(sf::st_cast(x.multipoints[i, ], "POINT"))
  }

  return(x.points)
}

osm_park_road_interesect <- osm_parks %>%
  st_boundary() %>%
  st_intersection(osm_roads) %>%
  select(name, size_ha, size_log) %>%
  st_multipoint_to_point()

# E. Combine both into final access point layer
park_accesspoints <- rbind(osm_park_centroids, osm_park_road_interesect)
```

### 2. Preparing Observer Locations

To represent potential origins of park visitors, we use a **10 m resolution grid** covering the entire area of interest. Points located **within parks** inherently have optimal accessibility, thus are excluded from further computation. Similarly, we remove points in water bodies.

```{r eval=FALSE}
# --- 2. Prepare Observer Locations ---

# A. Create a regular grid (10m resolution)
observer <- st_make_grid(aoi, cellsize = 50, what = "centers") %>%
  st_as_sf() %>%
  st_set_geometry("geom")
observer <- observer[aoi,]
observer <- observer %>%
  mutate(id = 1:n()) %>%
  relocate(id, .after = last_col())

# B. Remove points inside parks
obs_in_parks <- observer[osm_parks, ]
observer <- observer[-obs_in_parks$id, ]

# C. Remove points in water bodies
observer <- observer[-(observer[water_mask, ]$id), ]
```

### 3. Calculating Network-Based Distance Using OSRM

We use the **Open Source Routing Machine (OSRM)** to derive walking distances via a **road network**. OSRM offers **fast and efficient** routing by pre-processing large OSM data sets and applying advanced routing algorithms (e.g., Multi-Level Dijkstra).

Below is an example workflow for **downloading** the relevant OSM `.pbf` (protocol buffer) file (here, the Mittelfranken region of Germany) and **running OSRM** in a **Docker** container. Ensure Docker is installed beforehand ([docs.docker.com/engine/install/ubuntu](https://docs.docker.com/engine/install/ubuntu)).

```{r eval=FALSE}
# Download OSM data for your region of interest
if(!file.exists("01_analysis/0101_data/osrm/mittelfranken-latest.osm.pbf")){
  dir.create("01_analysis/0101_data/osrm/", showWarnings = FALSE)
  default_timeout <- getOption("timeout")
  options(timeout = max(600, default_timeout))
  download.file(
    url = "https://download.geofabrik.de/europe/germany/bayern/mittelfranken-latest.osm.pbf",
    destfile = "01_analysis/0101_data/osrm/mittelfranken-latest.osm.pbf",
    mode="wb"
  )
  options(timeout = default_timeout)
}

# OSRM Docker commands:
# 1) Navigate to the appropriate folder:
#    cd 01_analysis/0101_data/osrm/
# 2) Extract, partition, and customize:
#    sudo docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-extract -p /opt/foot.lua /data/mittelfranken-latest.osm.pbf
#    sudo docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-partition /data/mittelfranken-latest.osrm
#    sudo docker run -t -v "${PWD}:/data" osrm/osrm-backend osrm-customize /data/mittelfranken-latest.osrm
# 3) Run the OSRM server:
#    sudo docker run -t -i -p 5000:5000 -v "${PWD}:/data" osrm/osrm-backend osrm-routed --max-table-size=50000 --algorithm mld /data/mittelfranken-latest.osrm
```

After setting up OSRM, we retrieve walking distances for each **observer location** to the **k nearest** park access points, and compute the walking distance or time. These distance values are subsequently **weighted** by park size, recognizing that larger parks may attract users from farther afield [@giles-corti2005].

```{r eval=FALSE}
# --- 3. Network Analysis with OSRM ---

# The OSRM server is run locally via Docker (setup described earlier).

observer_park_duration <- function(j){
  # Identify k-nearest access points
  knn_access <- nngeo::st_nn(
    observer[j,],
    park_accesspoints,
    k = 2 * round(sqrt(nrow(park_accesspoints))),
    progress = FALSE
  ) %>% unlist()

  # Compute travel times using OSRM
  duration_table <- osrmTable(
    src = observer[j,],
    dst = park_accesspoints[knn_access,],
    osrm.server = "http://0.0.0.0:5000/",
    osrm.profile = "foot"
  )

  # Select best match (close & large)
  out <- tibble(
    id = observer[j,]$id,
    dist = as.numeric(duration_table$durations),
    size_ha = park_accesspoints[knn_access,]$size_ha,
    size_log = park_accesspoints[knn_access,]$size_log,
    name = park_accesspoints[knn_access,]$name
  ) %>%
    group_by(name) %>%
    filter(dist == min(dist)) %>%
    distinct() %>%
    ungroup() %>%
    # Normalize distance & size values
    mutate(dist_01 = dist / mean(dist) * 100,
           size_01 = size_log / mean(size_log) * 100) %>%
    mutate(dist_01 = 1 - dist_01 / max(dist_01),
           size_01 = size_01 / max(size_01)) %>%
    # Weight by distance vs. park size
    mutate(w = (1.91 / 0.85 * dist_01 + size_01) / 2) %>%
    filter(w == max(w)) %>%
    select(id, name, dist, size_ha, w)

  return(out)
}

# Split observer set into manageable chunks
obs_seq <- split(1:nrow(observer), ceiling(seq_along(1:nrow(observer))/250))

# Initialize an output CSV
tibble(
  id = as.numeric(),
  name = as.character(),
  dist = as.numeric(),
  size_ha = as.numeric(),
  w = as.numeric()
) %>%
  readr::write_csv("01_analysis/0101_data/osrm/access_table.csv", col_names = TRUE)

# Parallel loop over observer subsets
pb = pbmcapply::progressBar(min = 0, max = length(obs_seq))
stepi = 1
for(this_seq in obs_seq){
  dist_table <- parallel::mclapply(this_seq, observer_park_duration, mc.cores = 18) %>%
    do.call(rbind, .)

  readr::write_csv(dist_table, "01_analysis/0101_data/osrm/access_table.csv", append = TRUE)
  setTxtProgressBar(pb, stepi)
  stepi = stepi + 1
}
close(pb)

# Merge distance results with the observer SF
dist_table <- readr::read_csv("01_analysis/0101_data/osrm/access_table.csv")
dist_sf <- inner_join(observer, dist_table)

# Terminate OSRM server and cleanup if desired
unlink("01_analysis/0101_data/osrm/", recursive = TRUE)
```

### 4. Generating the GACI Raster

Finally, the weighted accessibility values (`w`) are **interpolated** into a continuous **10 m raster** using an **Inverse Distance Weighting (IDW)** approach, capping extreme values and **reclassifying** with **Jenks** to facilitate map visualization. Areas with lowest travel times to largest parks receive the most favorable accessibility scores.

```{r eval=FALSE}
# --- 4. Convert Access Values to GACI Raster ---

# A. Interpolate to 10m grid using IDW
gaci_raw <- CGEI::sf_interpolat_IDW(
  observer = dist_sf,
  v = "w",
  aoi = aoi,
  max_distance = Inf,
  n = 20,
  raster_res = 10,
  na_only = TRUE,
  cores = 22,
  progress = TRUE
)

# B. Park cells: assign ideal accessibility
park_cells <- terra::extract(gaci_raw, osm_parks, cells = TRUE, ID = FALSE)$cell
gaci_raw[park_cells] <- NA

# C. Reclassify via Jenks for readability
gaci <- CGEI:::reclassify_jenks(gaci_raw, n_classes = 9)
gaci[park_cells] <- 9

gaci <- gaci %>%
  crop(aoi) %>%
  mask(aoi)
names(gaci) <- "GACI"
```

```{r echo=FALSE}
gaci <- rast("../../../../CGEI_Nuernberg/01_analysis/0101_data/01_gaci.tif") %>% 
  crop(aoi) %>% 
  mask(aoi)
names(gaci) <- "GACI"
```

Below is the **final GACI map** for Nürnberg, indicating how readily residents can access public parks at each location. Areas in **red** reflect relatively low accessibility (longer walking distances to suitably large parks), whereas **green** signifies more favorable access. Unsurprisingly, the **city center** exhibits poorer scores, likely due to limited large parks and a denser built environment. By contrast, districts near the **eastern greenbelt** and forested zones (e.g., Reichswald) show stronger accessibility, thanks to extensive parkland and well-connected footpaths. Furthermore, the **Wöhrder Wiese** and **Wöhrder See,** as well as the **Pegnitz** in the west (close to the *Westfriedhof*) form a crucial green belt, ensuring many neighborhoods in densely populated areas benefit from relatively high accessibility.

Overall, the GACI reveals not just *where* parks are lacking, but also *how easily* citizens can reach them - an essential insight for guiding equitable urban planning and public health initiatives.

```{r echo=FALSE, fig.height=8, fig.width=8}
gaci_moegeldorf <- gaci %>% 
  crop(moegeldorf) %>% 
  mask(moegeldorf)

par(bg=NA)
mf_theme(bg = "transparent")

# 1. Main map
mf_map(aoi, expandBB = c(0, 0, 0, .35))
mf_raster(
    x = gaci, 
    type = "classes",
    nbreaks = 9,
    pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
    leg_pos = "bottomright",
    leg_frame = TRUE,
    add = TRUE
)

mf_map(aoi, add = TRUE, col = "transparent")
mf_map(moegeldorf, add = TRUE, col = "transparent", border = "blue", lwd = 2)

# 2. Inset
mf_inset_on(x = moegeldorf, pos = "topright", cex = .3)
mf_map(moegeldorf)
mf_raster(
    x = gaci, 
    type = "classes",
    nbreaks = 9,
    pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
    leg_pos = NA,
    add = TRUE
)
mf_map(aoi, add = TRUE)
mf_shadow(moegeldorf, add = TRUE)

mf_raster(
  x = gaci_moegeldorf, 
  type = "classes",
  nbreaks = 9,
  pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
  leg_pos = NA,
  add = TRUE
)

mf_map(aoi, add = TRUE, col = "transparent")

mf_title("Mögeldorf\nGACI",
  pos = "left", tab = TRUE,
  cex = 1.2, line = 2.5, inner = TRUE,
  bg = "gray80", fg = "black"
)
mf_scale(size = 1, cex = 0.8)
box()
mf_inset_off()

# 3. Credits and scale bar
mf_credits(txt = paste("Sources", 
                           "www.statistik.nuernberg.de", 
                           "Own calculations", 
                           sep = "\n"),
           cex = 0.8)
mf_scale(size = 5, cex = 0.8, pos = c(653880, 5466356))
```

## References
