---
title: Nürnberg - 1. Viewshed Greenness Visibility Index (VGVI)
author: ''
date: '2025-01-13'
slug: nuernberg-viewshed-greenness-visibility-index
categories: ["R"]
tags: ["Nürnberg", "VGVI", "R", "Greenness Visibility", "Greenspace", "CGEI"]
summary: "By combining binary greenspace data with DSM/DTM viewshed analysis, the Visibility-based Greenspace Index (VGVI) captures how much vegetation is actually seen from a ground-level perspective—offering a more user-centered and comprehensive measure of urban greenery."
image:
  focal_point: ""
  placement: 2
  preview_only: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The **visibility of greenspace** is crucial for understanding how urban environments influence mental health and well-being. In contrast to simple 2D buffer analyses, visibility focuses on the **eye-level perspective**: how much greenery a person sees from their usual vantage point. One way to estimate this is by using street-view (SV) images (e.g., from Google Street View or Baidu Street View). However, SV-based methods are often limited by seasonal inconsistencies and coverage gaps on roads inaccessible by car [@li2015]. A promising alternative is **viewshed analysis** within a GIS framework, which can accurately simulate what a person can see from specific locations. Recent studies at city-wide scales show this approach yields highly accurate visibility estimates, without depending on SV image availability [@tabrizian2020; @labib2021a_visibility; @cimburova2022]. Improved computation times now allow large-scale assessments of visible greenspace with relatively little effort [@brinkmann2022_agile]. This measure of greenspace visibility is especially relevant for the restoration pathway, where greenery supports mental recovery [@markevych2017].

Ample evidence suggests that **exposure to green spaces can significantly lower stress and restore cognitive function**. According to the stress reduction theory, natural environments promote stress relief by providing visual cues perceived as safe havens in evolutionary terms [@ulrich1981; @ulrich1991]. Meanwhile, the **attention restoration** theory posits that nature helps the mind recover from “attention fatigue” by engaging involuntary attention, thus giving directed attention a chance to rest [@kaplan1989]. Both theories underscore the importance of simply seeing natural elements for psychological benefits. As a result, the inclusion of visibility metrics - from either street-view imagery or GIS-based viewshed analysis - can add vital insights into how greenspace fosters restoration and supports better mental health outcomes [@dadvand2016].

## Data

The relevant spatial datasets have been uploaded on Zenodo: administrative boundaries (AOI), DTM and DSM rasters, and a binary land-cover raster for greenspace analysis. All raster data is available at 1m resolution.

```{r eval=FALSE}
library(sf)
library(dplyr)
library(terra)
library(CGEI)

zenodo_url <- "https://zenodo.org/records/14633167/files/"
# AOI
aoi <- read_sf(paste0(zenodo_url, "01_Nbg_Bezirke.gpkg")) %>%
  filter(code_bez != "97") %>% # This is not available in the GAVI
  group_by(code_bez, name_bez) %>%
  summarise() %>%
  ungroup()

# DSM
dsm <- rast(paste0(zenodo_url, "03_dsm_1m.tif"))

# DTM
dtm <- rast(paste0(zenodo_url, "03_dtm_1m.tif"))

# Greenspace (binary)
lulc <- rast(paste0(zenodo_url, "03_ndvi_01_1m.tif"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(terra)
library(CGEI)

zenodo_url <- "../../../../SpatData_Nbg/01_analysis/0102_data/02_processed/"

# AOI
aoi <- read_sf(paste0(zenodo_url, "01_Nbg_Bezirke.gpkg")) %>%
  filter(code_bez != "97") %>% # This is not available in the GAVI
  group_by(code_bez, name_bez) %>%
  summarise() %>%
  ungroup()

# DSM
dsm <- rast(paste0(zenodo_url, "03_dsm_1m.tif"))

# DTM
dtm <- rast(paste0(zenodo_url, "03_dtm_1m.tif"))

# Greenspace (binary)
lulc <- rast(paste0(zenodo_url, "03_ndvi_01_1m.tif"))
```

```{r echo=FALSE}
# Selection of a target district to display in the inset
moegeldorf <- aoi[aoi$code_bez == "92",]

moegeldorf_dsm <- dsm %>% 
  crop(moegeldorf) %>% 
  mask(moegeldorf) %>% 
  log()

moegeldorf_dtm <- dtm %>%
  crop(moegeldorf) %>%
  mask(moegeldorf)

moegeldorf_lulc <- lulc %>%
  crop(moegeldorf) %>%
  mask(moegeldorf)
```

```{r echo=FALSE, fig.height=8.5, fig.width=10}
library(mapsf)
dem_colors <- colorRampPalette(c(
  "#336600", "#81c31f", "#ffffcc", "#fade89", "#f4bd45", 
   "#95601E", "#70401B", "#66330c"
))(25)
dsm_colors <- colorRampPalette(c(
  "#336600", "#81c31f", "#ffffcc", "#fade89", "#f4bd45", 
   "#95601E", "#70401B", "#66330c"
))(75)

par(bg=NA)
mf_theme(bg = "transparent")

# 1. Main map
mf_map(aoi, expandBB = c(0, 0, 0, .35))
mf_map(moegeldorf, add = TRUE, col = "tomato")

# 2. Insets
# 2.1 DTM
mf_inset_on(x = moegeldorf, pos = "topright", cex = .3)
mf_map(moegeldorf)
mf_map(aoi, add = TRUE)
mf_shadow(moegeldorf, add = TRUE)

mf_raster(
  x = moegeldorf_dtm, 
  add = TRUE,
  pal = dem_colors, 
  breaks = seq(min(values(moegeldorf_dtm, na.rm = TRUE)), 
               max(values(moegeldorf_dtm, na.rm = TRUE)),
               length.out = 25),
  leg_pos = NA
)

mf_title("Mögeldorf\na) DTM",
  pos = "left", tab = TRUE,
  cex = 1.2, line = 2.5, inner = TRUE,
  bg = "gray80", fg = "black"
)
mf_scale(size = 1, cex = 0.8)
box()
mf_inset_off()

# 2.2 DSM
mf_inset_on(x = moegeldorf, pos = "right", cex = .3)
mf_map(moegeldorf)
mf_map(aoi, add = TRUE)
mf_shadow(moegeldorf, add = TRUE)

mf_raster(
  x = moegeldorf_dsm, 
  add = TRUE,
  pal = dem_colors, 
  breaks = seq(min(values(moegeldorf_dsm, na.rm = TRUE)), 
               max(values(moegeldorf_dsm, na.rm = TRUE)),
               length.out = 25),
  leg_pos = NA
)

mf_title("Mögeldorf\nb) DSM",
  pos = "left", tab = TRUE,
  cex = 1.2, line = 2.5, inner = TRUE,
  bg = "gray80", fg = "black"
)
mf_scale(size = 1, cex = 0.8)
box()
mf_inset_off()

# 2.3 LULC
mf_inset_on(x = moegeldorf, pos = "bottomright", cex = .3)
mf_map(moegeldorf)
mf_map(aoi, add = TRUE)
mf_shadow(moegeldorf, add = TRUE)

mf_raster(
  x = moegeldorf_lulc, 
  add = TRUE,
  pal = c("firebrick2", "green4"), 
  breaks = c(0, 1),
  leg_pos = NA
)

mf_title("Mögeldorf\nc) LULC",
  pos = "left", tab = TRUE,
  cex = 1.2, line = 2.5, inner = TRUE,
  bg = "gray80", fg = "black"
)
mf_scale(size = 1, cex = 0.8)
box()
mf_inset_off()

# 3. Credits and scale bar
mf_credits(txt = paste("Sources", 
                           "www.statistik.nuernberg.de", 
                           "geoportal.bayern.de", 
                           "Own calculations", 
                           sep = "\n"),
           cex = 0.8)
mf_scale(size = 5, cex = 0.8, pos = c(653880, 5466356))
```

Nürnberg, shown on the left in grey, is divided into 86 administrative districts. Mögeldorf has been highlighted in orange to provide examples of the three key greenness metrics on the right:

1.  **DTM** - The Digital Terrain Model illustrates the bare earth’s surface elevation, excluding buildings and vegetation. In the map, higher terrain is depicted in brownish colors, transitioning to green in lower-lying areas.

2.  **DSM** - The Digital Surface Model includes all features above ground (e.g., buildings, trees). The green and yellow shades indicate varying heights, reflecting how built-up areas and vegetation rise above the terrain.

3.  **LULC** - The land-use/land-cover map distinguishes vegetated (green) from non-vegetated (red) areas, making it easy to spot where greenery dominates versus dense urban structures.

## VGVI

A key measure of **eye-level** greenspace exposure is the Viewshed Greenness Visibility Index (**VGVI**) [@labib2021a_visibility]. This metric quantifies how much greenery (based on a binary LULC layer) is visible from a given point when accounting for local surface elevations (DSM) and terrain elevations (DTM).

### 1. Defining Observer Locations

To simulate where an individual might stand, we first identify valid observer points at a 5 m resolution, excluding areas where buildings or other tall structures exceed the observer’s eye height (e.g., 2.2 m above ground). The code below shows how we aggregate DSM and DTM to 5 m, filter for valid points `(dsm5 <= dtm5 + 2.2)`, and convert these locations into an SF object:

```{r eval=FALSE}
# First we need all valid cells in a 5x5m grid. These are where the DSM is not
# higher than 1.8m + DTM
dsm5 <- aggregate(dsm, 5)
dtm5 <- aggregate(dtm, 5)

obs <- dsm5 <= (dtm5 + 2.2)
obs <- obs %>%
  crop(aoi) %>%
  mask(aoi)

obs_vals <- values(obs)
obs_vals <- which(as.vector(obs_vals))

# Convert the cell coordinates as an SF
obs_sf <- terra::xyFromCell(obs, obs_vals)
obs_sf <- st_as_sf(as_tibble(obs_sf), coords = c("x", "y"), crs = crs(dsm)) %>%
  mutate(id = 1:n()) %>%
  relocate(id)
```

### 2. Calculating the VGVI

With all potential observer locations defined, we compute the **VGVI** using a **viewshed** approach. In brief:

-   The observer’s eye height (2.2 m) is placed atop the DTM, and any obstacles (e.g., buildings, trees) in the DSM are considered.

-   Each visible line-of-sight cell in a viewshed is checked against the binary greenspace raster (LULC).

-   A distance decay function reduces the weight of objects further away (e.g., **exponential** with m=1m and b=3).

Below is the `vgvi()` function call from the [**CGEI**](https://github.com/STBrinkmann/CGEI/) R package that performs these steps in parallel:

```{r eval=FALSE}
# Calculate VGVI
vgvi_sf <- vgvi(observer = obs_sf,
                dsm_rast = dsm, dtm_rast = dtm, greenspace_rast = lulc,
                max_distance = 500, observer_height = 2.2, 
                m = 1, b = 3, mode = "exponential",
                cores = 22, progress = TRUE)
```

### 3. Interpolating VGVI to a Continuous Surface

Finally, the point-based VGVI results (one value per observer location) are interpolated to a continuous 10 m resolution raster using **Inverse Distance Weighting (IDW)**. IDW assumes nearby values are more similar than distant ones, so each unknown cell’s VGVI is a distance-weighted average of its nearest neighbors [@hartmann2018]. Relevant parameters (e.g., number of neighbors, distance threshold) can be fine-tuned based on sensitivity analysis.

```{r eval=FALSE}
# Now rasterize the VGVI
vgvi_sf <- vgvi_sf[aoi,]
vgvi_rast_05 <- sf_interpolat_IDW(observer = vgvi_sf,
                                  v = "VGVI",
                                  aoi = aoi,
                                  raster_res = 5,
                                  n = 10, beta = 2, max_distance = 500,
                                  na_only = TRUE,
                                  cores = 22, progress = TRUE)

# Bring to 10m resolution
vgvi_rast_10 <- aggregate(vgvi_rast_05, 2)
vgvi_rast_10 <- vgvi_rast_10 %>%
  crop(aoi) %>%
  mask(aoi)
```

Finally, the **VGVI** raster can be classified using **Jenks natural breaks** to highlight different visibility classes.

```{r eval=FALSE}
# Classify the VGVI using Jenks natural breaks - N=9
vgvi_rast_10 <- CGEI:::reclassify_jenks(vgvi_rast_10, 9)
vgvi_rast_10 <- as.int(vgvi_rast_10)
names(vgvi_rast_10) <- "VGVI"
```

```{r echo=FALSE}
vgvi_rast_10 <- rast("../../../../CGEI_Nuernberg/01_analysis/0101_data/01_vgvi.tif") %>% 
  crop(aoi) %>% 
  mask(aoi)
names(vgvi_rast_10) <- "VGVI"
```

Below is the **final VGVI map** for Nürnberg, showing how much greenspace is visible from an eye-level perspective at each location. Areas in **red** represent lower greenness visibility, whereas **green** areas indicate higher greenness visibility. As expected, more **densely built-up** districts near the city center display lower VGVI values, while **peripheral areas** - which often have less building coverage and more open space - reveal higher VGVI values.

Overall, this view-dependent approach highlights not only *where* greenspace is located, but also *how much* of it is actually visible to an individual walking through the city - critical information for urban design, mental health research, and ecological planning.

```{r echo=FALSE, fig.height=8, fig.width=8}
vgvi_moegeldorf <- vgvi_rast_10 %>% 
  crop(moegeldorf) %>% 
  mask(moegeldorf)

par(bg=NA)
mf_theme(bg = "transparent")

# 1. Main map
mf_map(aoi, expandBB = c(0, 0, 0, .35))
mf_raster(
    x = vgvi_rast_10, 
    type = "classes",
    nbreaks = 9,
    pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
    leg_pos = "bottomright",
    leg_frame = TRUE,
    add = TRUE
)

mf_map(aoi, add = TRUE, col = "transparent")
mf_map(moegeldorf, add = TRUE, col = "transparent", border = "blue", lwd = 2)

# 2. Inset
mf_inset_on(x = moegeldorf, pos = "topright", cex = .3)
mf_map(moegeldorf)
mf_raster(
    x = vgvi_rast_10, 
    type = "classes",
    nbreaks = 9,
    pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
    leg_pos = NA,
    add = TRUE
)
mf_map(aoi, add = TRUE)
mf_shadow(moegeldorf, add = TRUE)

mf_raster(
  x = vgvi_moegeldorf, 
  type = "classes",
  nbreaks = 9,
  pal = RColorBrewer::brewer.pal(9, "RdYlGn"),
  leg_pos = NA,
  add = TRUE
)

mf_map(aoi, add = TRUE, col = "transparent")

mf_title("Mögeldorf\nGAVI",
  pos = "left", tab = TRUE,
  cex = 1.2, line = 2.5, inner = TRUE,
  bg = "gray80", fg = "black"
)
mf_scale(size = 1, cex = 0.8)
box()
mf_inset_off()

# 3. Credits and scale bar
mf_credits(txt = paste("Sources", 
                           "www.statistik.nuernberg.de", 
                           "geoportal.bayern.de", 
                           "Own calculations", 
                           sep = "\n"),
           cex = 0.8)
mf_scale(size = 5, cex = 0.8, pos = c(653880, 5466356))
```

## References

